---
title: "Assignment 3 - Data Management and Programming for Epidemiology 401179"
date: "7/06/2016"
output: pdf_document
authors: Kevin Zhu, Henry Tangi, Pronati Das
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Assignment 3 involves manipulating some large, real-life data files. These files contain publicly-available data on adverse drug reactions compiled by the US FDA (Food and Drug Administration). Four data files will be used, containing data submitted to the FDA in the 1st quarter of 2014. The four files can be found in the Assignment 3 directory on the course web site on vUWS. You should create a new RStudio project for the assignment, and unzip the data files in the project directory so they are accessible to the project.

Information on the content of each of the data files can be found on the following web pages:

 * Demographics: http://www.nber.org/fda/faers/2014/desc/demo2014q1/desc.txt
 * Drug information http://www.nber.org/fda/faers/2014/desc/drug2014q1/desc.txt
 * Adverse reaction description http://www.nber.org/fda/faers/2014/desc/reac2014q1/desc.txt
 * Outcomes http://www.nber.org/fda/faers/2014/desc/outc2014q1/desc.txt

Your solution needs to be submitted as an Rmd (R Markdown) file which will be used to create an HTML (web) document combining the descriptions of the assignment tasks with your R code solutions, and the output created by that R code.

All data manipulation and table and chart preparation must be done by R code contained in your solution file. Do not manipulate data using other means - marks will be deducted if you do. The goal is to create a 100% reproducible analysis.

Use this Rmd file as a template for your solution (feel free to add relevant text or commentary). You do not need to submit copies of the data files - the lecturers have copies of those. Your solution will be marked by running your Rmd solution file using knitr in RStudio. Submit it as a file named Assignment3_solution_name1_name2_name3.Rmd, where name1, name2 and name3 are the surnames of the students collaborating that that solution file. Up to three students may collaborate on one solution. Thus your solution should only rely on the the four data files supplied. You may use any additional R packages or libraries which you like - the lecturers will install these additional packages on their own system as required in order to run and thus mark your solution. The use of the dplyr and ggplot2 packages is recommended.

Task a. in Section 1 below has been done for you to show you how to include your R code in this file. You may use any additional knitr directives or options that you wish, but please leave the value of echo at the default of TRUE so that your code is visible in the output HTML documemt.

# Due Date

Midnight 5th June 2016.

# Statement of contributions

Kevin Zhu - everything

# Section 1

## Value

7 marks

## Tasks

a. Read the demographics file for qtr1 2014 into a data frame in R, but do not automatically convert strings to factors

```{r}
Demographics <- read.csv('demo2014q1.csv', stringsAsFactors = FALSE)
```

b. Convert the i_f_code column into a factor with appropriate labels.

```{r}
Demographics$i_f_code <- factor(Demographics$i_f_code, levels = c("F", "I"), labels = c("Follow-up", "Initial"))
```

c. Convert the event_dt variable into a date variable using as.Date() and name the new column event_date. 

```{r}
Demographics$event_dt_factor <- factor(Demographics$event_dt)
Demographics$event_date <- as.Date(Demographics$event_dt_factor, "%Y%m%d")
```

d. Write a function to create a numeric age_yrs column in the demography data frame which has age in years. Create a column called agr_grp which contains the 5 year age group, from 0-4 through to 85+, for each record

```{r}
# For this solution, dplyr package needs to be installed
library("dplyr", lib.loc="~/R/win-library/3.2")

#Decade age converted
Decade_age <- Demographics %>%
filter(age > 0, age_cod == "DEC") %>%
mutate(decade_age_converted = age*10)
Demographics <- merge(Demographics, Decade_age,
all.x=TRUE)

#Month age converted
Month_age <- Demographics %>%
filter(age > 0, age_cod == "MON") %>%
mutate(month_age_converted = age/12)
Demographics <- merge(Demographics, Month_age,
all.x=TRUE)

#Week age converted
Week_age <- Demographics %>%
filter(age > 0, age_cod == "WK") %>%
mutate(week_age_converted = age/52)
Demographics <- merge(Demographics, Week_age,
all.x=TRUE)

#Day age converted
Day_age <- Demographics %>% 
filter(age > 0, age_cod == "DY") %>%
mutate(day_age_converted = age/365)
Demographics <- merge(Demographics, Day_age, all.x=TRUE)

#Hour age converted
Hour_age <- Demographics %>%
filter(age > 0, age_cod == "HR") %>%
mutate(hour_age_converted = age/8760)
Demographics <- merge(Demographics, Hour_age,
all.x=TRUE)

#Normal years
Year_age <- Demographics %>%
filter(age > 0, age_cod == "YR") %>%
mutate(year_age_converted = age)
Demographics <- merge(Demographics, Year_age, all.x=TRUE)

#Adding everything into one column
Demographics$age_yrs <- rowSums(Demographics[,c("decade_age_converted", "year_age_converted", "month_age_converted", "week_age_converted", "day_age_converted", "hour_age_converted")], na.rm = TRUE)

#Making ages into groups
Demographics$age_grp <- cut(Demographics$age_yrs,breaks=c(0,5,11,16,21,26, 31, 36, 41, 46, 51, 56, 61, 66, 71, 76, 81, 85, Inf),labels=c("0 to 4 yrs","5 to 10 yrs","11 to 15 yrs","16 to 20 yrs","21 to 25 yrs","26 to 30 yrs", "31 to 35 yrs", "36 to 40 yrs", "41 yrs to 45 yrs", "46 to 50 yrs", "51 to 55 yrs", "56 to 60 yrs", "61 to 65 yrs", "66 to 70 yrs", "71 to 75 yrs", "76 to 80 yrs", "81 to 84 yrs", "85+ yrs"), right=FALSE)
```

e. Convert sex to a factor with appropriate labels

```{r}
Demographics$sex <- factor(Demographics$sex, levels = c("UNK", "M", "F", "NS"), labels = c("Unknown", "Male", "Female", "Not Specified"))
```

f. Convert the wtstr colum to a numeric weight in kilograms and store in a column called weight

```{r}

#Need to convert the strings into numeric. NAs will be introduced
Demographics$weight_as_numeric <- as.numeric(Demographics$wtstr)

#Now for the calculations
Pound_to_kg <- Demographics %>%
mutate(pound_to_kg = weight_as_numeric * 0.453592) %>%
filter(weight_as_numeric > 0, wt_cod == "LBS") 
Demographics <- merge(Demographics, Pound_to_kg,
all.x=TRUE)

Gram_to_kg <- Demographics %>%
mutate(gram_to_kg = weight_as_numeric * 100) %>%
filter(weight_as_numeric > 0, wt_cod == "GMS") 
Demographics <- merge(Demographics, Gram_to_kg,
all.x=TRUE)

Kg_to_kg <- Demographics %>%
mutate(kg_to_kg = weight_as_numeric) %>%
filter(weight_as_numeric >0, wt_cod == "KG")
Demographics <- merge(Demographics, Kg_to_kg,
all.x=TRUE)

#Into a column called weight

Demographics$weight <- rowSums(Demographics[,c("kg_to_kg", "gram_to_kg", "pound_to_kg")], na.rm = TRUE)

```

g. Convert occp_code to a factor and store as a column named occupation

```{r}
Occupation_code_factor <- factor(Demographics$occp_cod)
Demographics$occupation <- Occupation_code_factor

```

h. Remove all columns except the following from the demographics data frame:
 * primaryid
 * caseid
 * caseversion
 * i_f_code
 * event_date
 * age_yrs
 * age_grp
 * sex
 * weight
 * occupation

```{r}
keep <- c("primaryid", "caseid", "caseversion", "i_f_code", "event_date", "age_yrs", "age_grp", "sex", "weight", "occupation")
Demographics <- Demographics[ , keep, drop = FALSE]
```

i. keep only the initial report for each case in the demographics data frame

```{r}
No_initial_report <- filter(Demographics, i_f_code != "Initial")
Demographics <- anti_join(Demographics, No_initial_report, by = "i_f_code")

```

j. demonstrate that all of the above steps in Section 1 have been executed correctly.

# Section 2

## Value

6 marks

## Tasks

a. read the drug2014q1.csv data file into a data frame, do not convert strings into factor automatically

```{r}
Drug <- read.csv('drug2014q1.csv', stringsAsFactors = FALSE)
```

b. convert the role_cod column to a factor with appropriate labels:

```{r}
Drug$role_cod <- factor(Drug$role_cod, levels = c("PS", "SS", "C", "I"), labels = c("Primary suspect drug", "Secondary suspect drug", "Concomitant", "Interacting"))
```

c. convert the route column to a factor to save space

```{r}
Drug$route <- factor(Drug$route)
```

d. convert the dechal column to a factor with appropriate labels

```{r}
Drug$dechal <- factor(Drug$dechal, levels = c("Y", "N", "U", "D"), labels = c("Positive dechallenge", "Negative dechallenge", "Unknown", "Does Not Apply"))
```

e. convert the rechal column to a factor with appropriate labels

```{r}
Drug$rechal <- factor(Drug$rechal, levels = c("Y", "N", "U", "D"), labels = c("Positive dechallenge", "Negative dechallenge", "Unknown", "Does Not Apply"))
```

f. Remove all columns except the following from the drugs data frame:
 * primaryid
 * caseid
 * drug_seq
 * role_cod
 * drugname
 * route
 * dechal
 * rechal

```{r}
keep_drug <- c("primaryid", "caseid", "drug_seq", "role_cod", "drugname", "route", "dechal", "rechal")
Drug <- Drug[ , keep_drug, drop = FALSE]
```

g. demonstrate that all of the above steps in Section 2 have been executed correctly.

# Section 3

## Value

2 marks

## Tasks

a. read the outc2014q1.csv data file into a data frame, do not convert strings into factor automatically

```{r}
Outcome <- read.csv('outc2014q1.csv', stringsAsFactors = FALSE)
```

b. convert the outc_code column to a factor with appropriate labels

```{r}
Outcome$outc_cod <- factor(Outcome$outc_cod, levels = c("DE", "LT", "HO","DS", "CA", "RI", "OT"), labels = c("Death", "Life-Threatening", "Hospitalisation Initial or Prolonged", "Disability", "Congenital Anomaly", "Required Intervention to prevent", "Other Serious Important Medical Event"))
```

# Section 4

## Value

2 marks

## Tasks

a. read the reac2014q1.csv data file into a data frame, convert strings into factor automatically

```{r}
Reaction <- read.csv('reac2014q1.csv', stringsAsFactors = TRUE)
```

b. calculate the total size in memory of the four data frames: demographics, drugs, outcomes, reactions

Hint: use the object.size() function

```{r}
#First, individually calculate the sizes of each of the data frames.
Demographic_size <- object.size(Demographics)
Drug_size <- object.size(Drug)
Reaction_size <- object.size(Reaction)
Outcome_size <- object.size(Outcome)
#Then combine them to find the total size in bytes.
Total_size <- Demographic_size + Drug_size + Reaction_size + Outcome_size
Total_size
```

# Section 5

## Value

6 marks

## Task

Present tables which show the number of missing values in each variable in each of the 4 data frames.

```{r}
#Demographics table total missing in a table
sapply(Demographics, function(x) sum(is.na(x)))
#Drug table total missing in a table
sapply(Drug, function(x) sum(is.na(x)))
#Reaction table total missing in a table
sapply(Reaction, function(x) sum(is.na(x)))
#Outcome table total missing in a table
sapply(Outcome, function(x) sum(is.na(x)))
```

# Section 6

## Value

6 marks

## Task

Find the 10 drugs (drug names) which most commonly cause adverse events, and create a table and a chart to show the age group distribution of people suffering adverse events caused by those drugs. Plot with and without missing age group. Use a different y-axis range for each drug. 

```{r}
#Need to make sure ggplot2 is active
library(ggplot2)
#Make everything in drugname lowercase and trimws
Drug$drugname <- tolower(Drug$drugname)
Drug$drugname <- trimws(Drug$drugname)
#Create a new dataframe with Drug and Demographics in it. We will lose some data which isn't attached to any Primary or Case IDs.
Demographics_drug <- left_join(Demographics, Drug)
#Now to find the top 10 drug names
Drug_frequency <-  Demographics_drug %>% group_by(drugname) %>% summarise(n=n()) %>% arrange(desc(n))
Top_10_drugs <- Drug_frequency[1:10,]$drugname
Top_10_drugs
#Frequencies of top 10 drugs
Top_10_frequency <- Demographics_drug %>% filter(drugname %in% Top_10_drugs) 
sort(table(Top_10_frequency$drugname), decreasing = T)

#The plot with labels with missing age groups
ggplot(Top_10_frequency, aes(x=Top_10_frequency$age_grp)) + geom_bar(width=.5) + facet_grid(drugname ~. , scales="free") + labs(title="Top 10 adverse reactions from drugs distributed with missing age groups", x="Age Groups", y="Frequency of adverse reactions")

#The plot with labels without missing age groups

Demographics_drug_frequency_no_missing <- select(Demographics_drug, drugname, age_grp) %>% 
                                      filter(!is.na(age_grp))
Demographics_drug_frequency_no_missing$age_grp <- as.character(Demographics_drug_frequency_no_missing$age_grp)

ggplot(Demographics_drug_frequency_no_missing, aes(x=Demographics_drug_frequency_no_missing$age_grp)) + geom_bar(width=.5) + facet_grid(drugname ~. , scales="free") + labs(title="Top 10 adverse reactions from drugs distributed without missing age", x="Age Groups", y="Frequency of adverse reactions") 

                    

```

# Section 7

## Value

6 marks

## Task

For each of the top 10 drugs, show the relative frequency of each type of outcome, in a table and in a chart. Draw a chart showing the missing category for outcome, and a second chart with the missing category excluded.

```{r}
#First, find the total frequency of drugs and their outcomes.
Drug_outcome <- left_join(Top_10_frequency, Outcome, by="primaryid")
Drug_outcome_frequency <- select(Drug_outcome, outc_cod, drugname)
table(Drug_outcome_frequency)
#Introduce drug outcome frequencies without NA's now. Only looking at complete cases.
Total_complete_rows <- sum(complete.cases(Drug_outcome_frequency))
#We can calculate our relative frequencies now.
Relative_frequency <- table(Drug_outcome_frequency) / (Total_complete_rows)
Relative_frequency
                             
#Our relative frequency in a graph 
Relative_frequency_dataframe <- data.frame(Relative_frequency)

ggplot(Relative_frequency_dataframe, aes(x=Relative_frequency_dataframe$drugname, y=Relative_frequency_dataframe$Freq)) + geom_bar(stat="identity") + facet_grid(outc_cod ~ ., scales = "free") + labs(title="Relative frequency of outcomes with the top 10 adverse drugs", x="Drug Name", y="Relative Frequency")

#Graph for missing category in outcome

ggplot(Drug_outcome_frequency, aes(x=Drug_outcome_frequency$outc_cod)) + geom_bar() + facet_grid(drugname ~ ., scales = "free") + labs(title="Total number of outcomes in drugs with missing values", x="Outcome", y="Number of cases")

#Graph with the missing category excluded

Drug_outcome_frequency_no_missing <- select(Drug_outcome_frequency, drugname, outc_cod) %>% 
                                      filter(!is.na(outc_cod))

ggplot(Drug_outcome_frequency_no_missing, aes(x=Drug_outcome_frequency_no_missing$outc_cod)) + geom_bar() + facet_grid(drugname ~ ., scales = "free") + labs(title="Total number of outcomes in drugs excluding missing values", x="Outcome", y="Number of cases")
```

# Section 8

## Value

6 marks

## Task

Find the 10 most common types of drug reaction, and using a function which you have written, for each type of reaction, present a table of the (up to) ten most common drugs that cause that type of reaction (show their frequencies as well).

```{r}
#Make everything in reaction is placed into lowercase and trimws'd to help group similar variables
Reaction$pt <- tolower(Reaction$pt)
Reaction$pt <- trimws(Reaction$pt)
#Then create a merged dataframe for Drug and Reaction
Drug_reaction <- left_join(Drug, Reaction)
#Now to find the most frequent drug reactions. This will be using the similar function from Q6
Reaction_list <- Drug_reaction %>% group_by(pt) %>% summarise(n = n()) %>% arrange(desc(n))
Top_10_drug_reaction<- Reaction_list[1:10,]$pt
Top_10_drug_reaction

Top_10_reaction_drug <- function(this_pt) {
     top10rxn <- Drug_reaction %>% filter(pt == this_pt) %>% group_by(drugname) %>% summarise(n = n()) %>% arrange(desc(n))
     return(top10rxn[1:10,])
}

library(knitr)
kable(Top_10_reaction_drug("drug ineffective"))
kable(Top_10_reaction_drug("injection site pain"))
kable(Top_10_reaction_drug("fatigue"))
kable(Top_10_reaction_drug("nausea"))
kable(Top_10_reaction_drug("pain"))
kable(Top_10_reaction_drug("diarrhoea"))
kable(Top_10_reaction_drug("headache"))
kable(Top_10_reaction_drug("vomiting"))
kable(Top_10_reaction_drug("dyspnoea"))
kable(Top_10_reaction_drug("arthralgia"))


```

# Section 9

## Value

9 marks

## Task

Present some interesting or informative aspect of the data that involves data items from more than one of the four data frames, in the form of at least one table and one chart (but no more than 6 tables and/or charts in total). Each group must present a different analysis for this section - marks will be deducted from all groups that present the same analysis as any other group. You will need to confer amongst groups to ensure there is no overlap. The lecturers are happy to advise via email on whether any proposed analyses are too similar, or to suggest some suitable analyses.

```{r}
#We will want to look at the frequency of the top 10 drugs that had adverse effects and see whether or not there is a difference in effect within gender. 
#We will use our Demographics_drug dataframe as it has the data we need, and attempt to build cross-frequency tables to see if there is a correlation. 

#Frequencies table for gender and top 10 adverse effect drugs
Frequency_table <- table(Top_10_frequency$sex, Top_10_frequency$drugname)

#Plotting a graph to see our results
ggplot(Top_10_frequency, aes(x=Top_10_frequency$sex)) + geom_bar() + facet_grid(drugname ~ ., scales = "free") + labs(title="Gender and top 10 adverse effect drugs", x="Gender", y="Number of cases")

#Although we have a lot of missing genders, it seems that the female proportion is quite high. We should look whether or not there is a correlation between being of female sex and having adverse results.
Female_demographic <- filter(Top_10_frequency, sex == "Female)

                            


```

